---
title: "Reds Technical Assignment"
author: "Billy Fryer"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)

```

```{r Question1, include = FALSE}
################## Packages

# Install Packages that aren't already installed
# Stolen From: https://statisticsglobe.com/r-install-missing-packages-automatically
my_packages <- c("tidyverse", "xgboost", "caret", "flextable", "ggthemes",
                 "ggtext", "scales")
not_installed <- my_packages[!(my_packages %in% installed.packages()[ , "Package"])]
if(length(not_installed)) install.packages(not_installed)

# Library Needed packages
library(tidyverse)
library(xgboost)
library(caret)
library(flextable)
library(ggthemes)

# Fixes a quick error message with flextable
set_flextable_defaults(fonts_ignore=TRUE)

####################### Read in data
raw_train <- read_csv("data/train.csv")
raw_test <- read_csv("data/test.csv")

# Quick test to see if pitchers are shared between the two data sets
test_pitchers <- raw_test$PITCHER_KEY %>% unique()
train_pitchers <- raw_train$PITCHER_KEY %>% unique()
test_pitchers %in% train_pitchers # All Falses
rm(test_pitchers, train_pitchers)
# No which means that I shouldn't include PITCHER_KEY in the model I create
# After looking in data dictionary, don't want to use UID either


selected_train <- raw_train %>% 
  # Get rid of those 2 variables out of the train data set
  select(-c(PITCHER_KEY, UID)) %>% 
  # Convert PITCH_TYPE_TRACKED_KEY to numeric
  mutate(PITCH_TYPE_TRACKED_KEY = case_when(
    PITCH_TYPE_TRACKED_KEY == 'SI' ~ 0,
    PITCH_TYPE_TRACKED_KEY == "FB" ~ 1,
    PITCH_TYPE_TRACKED_KEY == "CF" ~ 2,
    PITCH_TYPE_TRACKED_KEY == "CB" ~ 3,
    PITCH_TYPE_TRACKED_KEY == "SL" ~ 4,
    PITCH_TYPE_TRACKED_KEY == "CH" ~ 5))

# Separate data from label
data_variables <- selected_train %>% 
  select(-PITCH_TYPE_TRACKED_KEY) %>% 
  as.matrix()
data_label <- selected_train$PITCH_TYPE_TRACKED_KEY

# 60/40 Train/Test Split
n <- nrow(data_variables)
prop <- 0.6
set.seed(119)
train_index <- sample(n, size = n * prop)
# split train data and make xgb.DMatrix
train_data   <- data_variables[train_index,]
train_label  <- data_label[train_index]
train_matrix <- xgb.DMatrix(data = train_data,
                            label = train_label)
# split test data and make xgb.DMatrix
test_data  <- data_variables[-train_index,]
test_label <- data_label[-train_index]
test_matrix <- xgb.DMatrix(data = test_data, label = test_label)

############################### Modeling
# xgboost is a great a way to get reduce log loss big time
# Haven't done multiple classification xgboost before
# so I got help from https://rpubs.com/mharris/multiclass_xgboost
numberOfClasses <- unique(data_label) %>% length()
xgb_params <- list("objective" = "multi:softprob",
                   "eval_metric" = "mlogloss",
                   "num_class" = numberOfClasses)
nround    <- 50 # number of XGBoost rounds
cv.nfold  <- 5 # 5 Rounds of Cross Validation

# Fit cv.nfold * cv.nround XGB models and save OOF predictions
cv_model <- xgb.cv(params = xgb_params,
                   data = train_matrix,
                   nrounds = nround,
                   nfold = cv.nfold,
                   verbose = FALSE,
                   prediction = TRUE)

# Check out predictions
OOF_prediction <- data.frame(cv_model$pred) %>%
  # Compute which column has highest probability
  mutate(max_prob = max.col(., ties.method = "last"),
         # Train label + 1 is the pitch type
         # R doesn't do colmn x0 so it increases everything by 1
         label = train_label + 1) %>%
  # Convert back to pitch types
  mutate(max_prob = case_when(
    max_prob == 1 ~ 'SI',
    max_prob == 2 ~ "FB",
    max_prob == 3 ~ "CF",
    max_prob == 4 ~ "CB",
    max_prob == 5 ~ "SL",
    max_prob == 6 ~ "CH"),
    label = case_when(
      label == 1 ~ 'SI',
      label == 2 ~ "FB",
      label == 3 ~ "CF",
      label == 4 ~ "CB",
      label == 5 ~ "SL",
      label == 6 ~ "CH"),
    )

#head(OOF_prediction)

# Confusion matrix
confusionMatrix(factor(OOF_prediction$max_prob),
                factor(OOF_prediction$label),
                mode = "everything")
# Doesn't do the best with CF or SI, but mixes them up with FB
# So we can live with that I guess


########## Holdout Data
# Predict hold-out test set
bst_model <- xgb.train(params = xgb_params,
                       data = train_matrix,
                       nrounds = nround)
# Predict based on test_matrix
test_pred <- predict(bst_model, newdata = test_matrix)
test_prediction <- matrix(test_pred, nrow = numberOfClasses,
                          ncol=length(test_pred)/numberOfClasses) %>%
  t() %>%
  data.frame() %>%
  mutate(label = test_label + 1,
         max_prob = max.col(., "last")) %>%
  # Convert back to pitch types
  mutate(max_prob = case_when(
    max_prob == 1 ~ 'SI',
    max_prob == 2 ~ "FB",
    max_prob == 3 ~ "CF",
    max_prob == 4 ~ "CB",
    max_prob == 5 ~ "SL",
    max_prob == 6 ~ "CH"),
    label = case_when(
      label == 1 ~ 'SI',
      label == 2 ~ "FB",
      label == 3 ~ "CF",
      label == 4 ~ "CB",
      label == 5 ~ "SL",
      label == 6 ~ "CH"),
    )

# confusion matrix of test set
confusionMatrix(factor(test_prediction$max_prob),
                factor(test_prediction$label),
                mode = "everything")
# Same as before doesn't do the best with CF or SI
# but mixes them up with FB which we can live with


########## Predict raw_test data

# Make raw_test into proper matrices
# split train data and make xgb.DMatrix
final_test_matrix <- raw_test %>% 
  select(RELEASE_SIDE, SPIN_RATE_ABSOLUTE, RELEASE_HEIGHT, RELEASE_SPEED) %>% 
  as.matrix()

colnames(final_test_matrix)<- colnames(train_matrix)

test_pred <- predict(bst_model, newdata = final_test_matrix)
final_prediction <- matrix(test_pred, nrow = numberOfClasses,
                          ncol=length(test_pred)/numberOfClasses) %>%
  t() %>%
  data.frame()

names(final_prediction) <- c("SI", "FB", "CF", "CB", "SL", "CH")

output_csv <- bind_cols(UID = raw_test$UID, final_prediction) %>%
  # Reorder columns
  select(UID, CB, CF, CH, FB, SI, SL)

write.csv(x = output_csv,
          file = "data/solution.csv",
          row.names = FALSE)

# Clear environment for part 2
rm(list = ls())
```

```{r Part2-table1}
# Read in Test Data
test_data <- read_csv("data/test.csv")

# First thing I'm going to do is look at results by pitcher
# specifically where each pitcher generates strikes and bad contact
# Do this in percent
strikes_contact_by_pitcher <- test_data %>% 
  # Converted to character since it's a name
  group_by(PITCHER_KEY) %>% 
  # Summarize by contact type/strike result
  # Strikes / All No Contact
  summarise(S_RATE = sum(S, na.rm = TRUE) / n(),
            # Swinging Strikes / All Pitches not put in play
            SS_RATE = sum(SS, na.rm = TRUE) / n(),
            # Bad Contact / All Contact
            BAD_CONTACT_RATE = sum(CONTACT_QUALITY == 'bad', na.rm = TRUE) / 
              sum(!is.na(CONTACT_QUALITY))
            ) %>% 
  ungroup() %>% 
  # Convert Pitcher to Character
  mutate(PITCHER_KEY = as.character(PITCHER_KEY)) %>% 
  arrange(-SS_RATE)
  
# Table time
strikes_contact_by_pitcher %>% 
  
  flextable() %>% 
  align(align = "center",
        part = "all") %>% 
  set_header_labels(values = list(
    PITCHER_KEY = "Pitcher",
    S_RATE = "Strike Rate",
    SS_RATE = "Swining Strike Rate",
    BAD_CONTACT_RATE = "Poor Contact Rate")
    ) %>%
  # Color Columns and text
  bg(j = "S_RATE",
     bg = scales::col_numeric(palette = "Reds", 
                              domain = c(0.45, 0.51))) %>% 
  color(j = "S_RATE",
     color = "black") %>% 
  bg(j = "SS_RATE",
     bg = scales::col_numeric(palette = "Reds", 
                              domain = c(0.10, 0.15))) %>% 
  color(j = "SS_RATE",
     color = "black") %>%
  bg(j = "BAD_CONTACT_RATE",
     bg = scales::col_numeric(palette = "Reds", 
                              domain = c(0.55, 0.65))) %>% 
  color(j = "BAD_CONTACT_RATE",
     color = "black")%>%
  # Percent Formats
  set_formatter(S_RATE = function(x) sprintf("%.1f%%",  x*100)) %>% 
  set_formatter(SS_RATE = function(x) sprintf("%.1f%%", x*100)) %>% 
  set_formatter(BAD_CONTACT_RATE = function(x) sprintf("%.1f%%", x*100)) %>%
  colformat_int(j = "PITCHER_KEY")
```

```{r speed-spin}
ggplot(test_data, aes(x = RELEASE_SPEED,
                      y = SPIN_RATE_ABSOLUTE)) +
  geom_point() +
  scale_color_discrete() +
  facet_wrap(~PITCHER_KEY,
             nrow = 1) +
  labs(x = "Velocity",
       y = "Spin Rate",
       title = "Velocity by Spin Rate") +
  theme_foundation() +
  theme(axis.title = element_text(face = "bold"),
        panel.grid = element_blank(),
        plot.title = ggtext::element_markdown(hjust = 0.5,
                                              size = 14,
                                              face = "bold"),
        plot.subtitle = ggtext::element_markdown(hjust = 0.5,
                                                 size = 8,
                                                 face = "italic")
  )
```

\newpage

My preferred order of these Pitchers would be:

**1) Pitcher 90002**

Pitcher 90002 had the highest Strike Rate and Swinging Strike Rate out of the 5 pitchers. This helps make up for the fact that he ranked fourth in Poor Contact Rate (the percentage of times poor contact was made by the hitter compared to all the times contact was made). This was the overall difference between pitchers 90004 and 90002 for me. From the graph, his velocity is elite as well as enhanced even further by his change in velocity between the fastball and the off speed and breaking pitches.

**2) Pitcher 90004**

For me, comparing pitchers 90004 and 90002 was the toughest. Pitcher 9004 has the third highest Strike and Swinging Strike Rates. However, Pitcher 90004 does very well at drawing weak contact, the best of all the pitchers in the data set. He's good enough to be elusive to draw swings and misses/ get strikes called but even when contact is made, it's not strong contact. The difference for me ended up being that Pitcher 90002 consistently threw the ball harder

**3) Pitcher 90000**

Pitcher 90000 has the second highest Strike Rate of all the pitchers in the test set, but what makes him really interesting is that his Swinging Strike Rate is nearly equal to Pitcher 90002 while his fastest pitches are a little bit slower. Although when he is hit he is hit hard, his high breaking pitch helps keep hitters off balance enough for that not to be as big of an issue.

**4) Pitcher 90001**

The biggest thing going for Pitcher 90001 is that he has the second highest Poor Contact Rate out of all the pitchers in the data set. His Strike Rate is below 50% which means he throws more balls than strikes on balls not hit. His peak velocity isn't that high either, maxing out around 95.5 mph. Looking at his Velocity by Spin Rate plot, he appears to throw 4 different pitches.

**5) Pitcher 90003**

Pitcher 9003 was consistently in the lower ranks across all three metrics. In addition, while looking at his Velocity vs Spin Rate graph, there were a few pitches that spin rates really close to 0. These of course be due to mistakes in the data but if that's a real pitch such as a knuckle ball that could be interesting to look into further.